---
title: "HoloFoodR: a statistical programming framework for holo-omics data integration workflows"
date: "`r Sys.Date()`"
package: HoloFoodR
bibliography: ../HoloFoodR.bib
output:
   BiocStyle::html_document:
       fig_height: 7
       fig_width: 10
       toc: yes
       toc_depth: 2
       number_sections: true
vignette: >
  %\VignetteIndexEntry{Case study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Introduction

[HoloFood database](https://www.holofooddata.org) is a large collection of
holo-omic and multi-omic data from two animal systems, chicken and salmon. It
was created by the HoloFood consortium. One of its objectives is to study the
interactions between animal systems, their microbiomes, and feed additives
to optimize the diet strategies of these farm animals.

For make it easily accessible, the developers provide an Application Programming
Interface (API) that permits interaction with programming languages, such as R
or Python.

We improve the accessibility by creating the HoloFoodR package that simplifies
API interaction and assists translating raw database data into R/Bioconductor
data containers, connecting to a vast ecosystem of bioinformatics R packages.

We do not aim to demonstrate HoloFoodR in isolation from the rest of the
ecosystem, but to showcase the possibility of data integration from other
databases, such as MGnify, which holds metagenomic data. Furthermore, we
provide a workflow from data exploration to advanced machine learning and
multi-omics, offering a practical example for readers.

Our main study questions are:

* How does treatment influence the gut microbiota of salmon?
* Do gut flora and fatty acids composition evolve over time?
* Is there a relationship between gut microbiota and the fatty acid composition
in muscle tissue?

```{r}
#| label: start
#| include: false

library(knitr)
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    cache = TRUE
)
```

```{r}
#| label: setup
#| output: false

# List of packages that we need
packages <- c(
    "dplyr", "DT", "ggsignif", "HoloFoodR", "MGnifyR", "mia", "miaViz", "MOFA2",
    "patchwork", "reticulate", "scater", "shadowtext"
)

# Load all packages into session. Stop if there are packages that were not
# successfully loaded
pkgs_not_loaded <- !sapply(packages, function(pkg){
    suppressPackageStartupMessages(require(pkg, character.only = TRUE))
})
pkgs_not_loaded <- names(pkgs_not_loaded)[pkgs_not_loaded]
if (length(pkgs_not_loaded) > 0) {
    stop("Error in loading the following packages into the session: '",
        paste0(pkgs_not_loaded, collapse = "', '"), "'")
}
```

## Import data

We start the workflow from data retrieval. We will use
[salmon data](https://www.holofooddata.org/samples/?animal__system=salmon)
and its associated fatty acid and metagenomic amplicon as an example.

### Retrieve HoloFood data

First of all, we have to query the HoloFood database to retrieve the salmon
accession numbers.

```{r}
#| label: get_animals
#| eval: false

# Get salmon samples
salmons <- HoloFoodR::doQuery("animals", system = "salmon", use.cache = TRUE)

# Get only the data that has both metagenomic amplicon and fatty acid
# data
salmons <- salmons |>
    filter(fatty_acids == TRUE & metagenomic_amplicon == TRUE)

colnames(salmons)
```

Next, we can retrieve the data associated with each salmon.

```{r}
#| label: get_animal_data
#| eval: false

# Get salmon data
salmon_data <- HoloFoodR::getData(
    accession.type = "animals",
    accession = salmons[["accession"]],
    use.cache = TRUE
)

# Get salmon samples
salmon_samples <- salmon_data[["samples"]]

# Get sample IDs
salmon_sample_ids <- unique(salmon_samples[["accession"]])

head(salmon_sample_ids)
```

The data returned above is a list of all sample accession numbers that are
associated with all salmons. For example, metagenomic amplicon samples, such as
[SAMEA112750580](https://www.holofooddata.org/sample/SAMEA112750580)
or fatty acid samples,
[SAMEA112950027](https://www.holofooddata.org/sample/SAMEA112950027).

We can use these accession numbers to fetch the data associated with each
sample type and store them as `experiments` in a `MultiAssayExperiment` (`MAE`)
object.

```{r}
#| label: get_salmon_mae
#| eval: false

# Get salmon <- experiments as MAE object
mae <- HoloFoodR::getResult(
    salmon_sample_ids,
    use.cache = TRUE
)
```

```{r}
#| label: save_salmon_mae
#| echo: false
#| eval: false

# Save salmon MAE This must be named differently
path <- system.file(
    "extdata", "salmon_mae_without_mgnify.RDS", package = "HoloFoodR")
saveRDS(object = mae, file = path)
```

```{r}
#| label: read_salmon_mae
#| echo: false

# Read MAE to avoid retrieval
path <- system.file(
    "extdata", "salmon_mae_without_mgnify.RDS", package = "HoloFoodR")
mae <- readRDS(file = path)
```

### Fetch metagenomic data from MGnify

HoloFood database does not include the data for metagenomic data. This data
can be retrieved from the [MGnify portal](https://www.ebi.ac.uk/metagenomics).
For this purpose, we will use
[MGnifyR package](https://bioconductor.org/packages/release/bioc/html/MGnifyR.html),
which in a similar fashion to HoloFoodR, allows simple interaction with MGnify
API.

```{r}
#| label: get_metagenomic_samples
#| eval: false

# Create MGnify object
mg <- MgnifyClient(
    useCache = TRUE,
    cacheDir = ".MGnifyR_cache"
)

# Select only metagenomic_amplicon sample type
metagenomic_salmon_samples <- salmon_samples |>
    filter(sample_type == "metagenomic_amplicon")

# Search for sample IDs in MGnify database
salmon_analysis_ids <- searchAnalysis(
    mg,
    type = "samples",
    metagenomic_salmon_samples[["accession"]]
)
```

`salmon_analysis_ids` character vector holds associations of HoloFood
metagenomic amplicon accession numbers (SAMEAxxxxxx) to their counterparts in
MGnify database (MGYAxxxxxx).

```{r}
#| label: get_metagenomic
#| eval: false

# Get metagenomic taxonomic data for salmon from MGnify
tse <- MGnifyR::getResult(
    mg,
    accession = salmon_analysis_ids,
    get.func = FALSE
)
```

```{r}
#| label: save_metagenomic
#| eval: false
#| echo: false

# Save salmon metagenomic TreeSE
path <- system.file(
    "extdata", "salmon_metagenomic_tse.RDS", package = "HoloFoodR")
saveRDS(object = mae, file = path)
```


```{r}
#| label: read_metagenomic
#| echo: false

# Read in salmon metagenomic TreeSE object
path <- system.file(
    "extdata", "salmon_metagenomic_tse.RDS", package = "HoloFoodR")
tse <- readRDS(file = path)
```

Data fetched from MGnify has MGnify-specific identifiers. We have to first
rename samples with HoloFood specific ID and then add the data to
`MultiAssayExperiment` combining all the data.

```{r}
#| label: add_metagenomic_data

# Add MGnify results to HoloFood data
mae <- addMGnify(tse, mae)
```

Now we have retrieved all the data that we are interested in this workflow.

## Data preprocess

Data cleaning is one of the most time-consuming and most important steps in
data analysis. For instance, we need to handle missing data, transform data
assays, and agglomerate the data.

In the next steps, we will:

1. Filter data
2. Ensure that the data is in correct format for analysis
3. Agglomerate data
4. Transform data

### Wrangle the data

Below we see upset plot that summarizes the available experiments and how
samples overlap between them.

```{r}
#| label: upsetplot
#| fig-cap: The distribution of experiments in the dataset, along with the number of samples in each experiment.

upset_plot <- upsetSamples(mae)
upset_plot
```

For demonstration purposes, we will focus on investigating fatty acids and
metagenomic data within the trial A performed by the HoloFood consortium. This
trial the health effects of fermented seaweed added to the diet of salmons.
The following script subsets the data to include aforementioned data.

```{r}
#| label: preprocess1

# Harmonize experiment names
names(mae) <- names(mae) |>
    tolower() |>
    gsub(pattern = " ", replacement = "_")

# Fetch only experiments that we need
mae <- mae[, , c("fatty_acids_mg", "metagenomic")]
names(mae) <- c("fatty_acids", "metagenomic")
# Filter MAE object to include only Trial A
mae <- mae[, colData(mae)[["Trial code"]] == "SA", ]
```

Some values of fatty acids are under detection thresholds. We assume them to be
zeroes. Moreover, the data includes a feature that just states from where the
fatty acids were collected. We remove this feature to ensure that the assay
contains only numeric values.

```{r}
#| label: preprocess2

# From metabolomic data, remove organ-fatty acids row because it only contains a
# string value "muscle" which denotes where the sample was drawn from
tse <- mae[[1]]
tse <- tse[!(rowData(tse)[["marker.name"]] %in% c("Organ-fatty acids")), ]
mae[[1]] <- tse

# Transform matrix to numeric. Some values are "< 0.01"
# If a number is < 0.01, assume it to be 0
assay <- assay(mae[[1]], "counts")
assay[assay == "<0.01"] <- 0
assay <- apply(assay, c(1, 2), function(x) as.numeric(gsub(",", ".", x)))

# Reassign assay back to MAE
assay(mae[[1]], "counts") <- assay
```

Moreover, we wrangle the sample metadata so that it includes all necessary
information.

```{r}
#| label: preprocess3

# Add time points
timepoints <- colData(mae[[2]])
timepoints <- timepoints[
    match(timepoints[["animal"]], rownames(colData(mae))), ]
timepoints <- ifelse(timepoints[["trial.timepoint"]] == 0, "start", "end")
timepoints <- factor(timepoints, levels = c("start", "end"))
colData(mae)[["timepoint"]] <- timepoints

# Add treatment groups
colData(mae)[["study_group"]] <- ifelse(
    colData(mae)[["Treatment concentration"]]>0, "treatment", "control")
colData(mae)[colData(mae)[["timepoint"]] == "start" , "study_group"] <-
    "control"

# Add animal metadata to separate experiments
mae[[1]] <- getWithColData(mae, 1)
mae[[2]] <- getWithColData(mae, 2)
```

### Filtering and agglomeration

Next, we can agglomerate features by prevalence to reduce the number of
low-abundant taxa and contaminants.

First, we visualize prevalence distribution of taxa with a histogram to decide
the prevalence threshold to use. We use 0.2% detection level to filter out
extremely low-abundant genera.

```{r}
#| label: prevalence_histgoram
#| fig-cap: The prevalence of microbial genera across samples.

# Add relative abundance data
mae[[2]] <- transformAssay(mae[[2]], method = "relabundance")

# Compute prevalence of relative abundance of microbial genera at detection
# level of 0.1%
prevalence <- getPrevalence(
    mae[[2]],
    rank = "Genus",
    assay.type = "relabundance",
    na.rm = TRUE,
    sort = TRUE,
    detection = 0.2 / 100
)

# Exclude microbes with 0 prevalence
prevalence <- prevalence[prevalence != 0]

hist(prevalence, main = "", xlab = "Prevalence")
```

We can also look at the raw prevalence numbers.

```{r}
#| label: display_prevalence_numbers

# Sort prevalence in decreasing order
sort(prevalence, decreasing = TRUE) |> head(10)
```

_Mycoplasma_ is present in all samples, which is not surprising as this genus
was found to be one of the most common in salmon intestine
(see @zarkasiPyrosequencingbasedCharacterizationGastrointestinal2014).

We then agglomerate our data by prevalence and by taxonomic rank to obtain group
all genera which are below the specified thresholds to the "Other" group.
This step is necessary to ensure that we only work with the most relevant
taxa, while not excluding the rest of data points. We use thresholds 20% and
0.2% for prevalence and detection, respectively.

```{r}
#| label: agglomerate_prevalence

# Agglomerate by prevalence by genus
altExp(mae[[2]], "prev_genus") <- agglomerateByPrevalence(
    mae[[2]],
    assay.type = "relabundance",
    rank = "Genus",
    prevalence = 20 / 100,
    detection = 0.2 / 100
)
```

Due to the limited number of samples, we also filter the fatty acid data to
include only those fatty acids that show variation within the dataset. The
rationale is that if a fatty acid does not vary, it cannot exhibit differences
between groups. We can find out a good threshold for the cutoff with a
histogram of standard deviations of fatty acid abundances.

```{r}
#| label: filter_fatty1
#| fig-cap: The distribution of standard deviations of fatty acid abundances.

rowData(mae[[1]])[["sd"]] <- rowSds(assay(mae[[1]], "counts"), na.rm = TRUE)
hist(rowData(mae[[1]])[["sd"]], breaks = 30, main = "",
    xlab = "Standard deviation")

# Increase the number of x axis ticks
x_labels <- seq(from = min(assay(mae[[1]])), to = max(assay(mae[[1]])), by = 1)
axis(side = 1, at = x_labels, labels = x_labels)
```

- Percentage of fatty acids with a standard deviation (SD) below 0.5:
`r round(mean(rowData(mae[[1]])[["sd"]] < 0.5), 3)*100`%
- Percentage of fatty acids with a standard deviation (SD) below 1:
`r round(mean(rowData(mae[[1]])[["sd"]] < 1), 3)*100`%

We apply a filtering threshold of 0.5 to exclude fatty acids that do not
exhibit sufficient variation in the dataset. Since most standard deviations
are below 1, it is a reasonable number if we do not want to exclude too many
fatty acids.

```{r}
#| label: filter_fatty2

mae[[1]] <- mae[[1]][ rowData(mae[[1]])[["sd"]] > 0.5, ]
```

For more detailed analysis, we pick certain fatty acids that have
well-established biological relevance. These include:

* Docosahexaenoic acid (DHA)
* Eicosapentaenoic acid (EPA)
* Alpha-linolenic acid
* Arachidonic acid
* Linoleic acid
* Oleic acid
* Palmitic acid
* Stearic acid

```{r}
#| label: filter_fatty3

relevant_fatty_acids <- c(
    "Docosahexaenoic acid 22:6n-3 (DHA)",
    "Eicosapentaenoic acid 20:5n-3 (EPA)",
    "Alpha-Linolenic acid 18:3n-3",
    "Arachidonic acid 20:4n-6 (ARA)",
    "Linoleic acid 18:2n-6",
    "Oleic acid 18:1n-9",
    "Palmitic acid 16:0",
    "Stearic acid 18:0"
)
altExp(mae[[1]], "relevant") <- mae[[1]][
    rownames(mae[[1]]) %in% relevant_fatty_acids, ]
```

### Transformation

We transform metagenomic counts with relative transformation and
centered log-ratio method to tackle the compositional data
(see @quinnFieldGuideCompositional2019).

```{r}
#| label: transformation_metagenomics

# Transform microbiome with centered log-ratio method
mae[[2]] <- transformAssay(
    mae[[2]],
    assay.type = "counts",
    method = "relabundance",
    MARGIN = "cols",
    altexp = altExpNames(mae[[2]])
)
mae[[2]] <- transformAssay(
    mae[[2]],
    assay.type = "counts",
    method = "clr",
    pseudocount = TRUE,
    MARGIN = "cols",
    altexp = altExpNames(mae[[2]])
)
```

Fatty acid data is already compositional as it is measured as concentration
(mg/g). We apply a log10 transformation to address
skewness in the data. Finally, the data is standardized to ensure all features
are on a comparable scale.

```{r}
#| label: transform_fatty

mae[[1]] <- transformAssay(
    mae[[1]],
    assay.type = "counts",
    method = "log10",
    MARGIN = "cols"
)
mae[[1]] <- transformAssay(
    mae[[1]],
    assay.type = "log10",
    method = "standardize",
    MARGIN = "rows"
)
```

## Analyzing fatty acids: Time and treatment effects

We aim to investigate how fatty acids evolve over time and, more importantly,
whether the feed additive impacts these fatty acids. To achieve this, we will
fit a simple linear model for each fatty acid, accounting for time and treatment
effects in a non-parametric manner. These models resemble the Wilcoxon test but
include additional covariates. This approach will allow us to estimate the
variability explained by time, treatment, or both factors.

```{r}
#| label: fatty_and_treatment

# Get fatty acids that we are going to test
tse <- altExp(mae[[1]], "relevant")

# For each fatty acid, fit linear model
res <- lapply(rownames(tse), function(feat){
    # Get data of single feature
    df <- meltSE(tse[feat, ], add.col = TRUE)
    # Fit model
    res <- lm(rank(counts) ~ study_group + timepoint, data = df)
    # Get only p-values
    res <- summary(res)
    res <- res[[4]][, 4]
    return(res)
})
# Combine results and adjust p-values
res <- do.call(rbind, res) |> as.data.frame()
res <- lapply(res, p.adjust, method = "fdr") |> as.data.frame()
# Add feature names
res[["feature"]] <- rownames(tse)
res |> datatable()
```

As observed, treatment does not appear to affect fatty acid concentrations
significantly. In contrast, time seems to influence nearly all fatty acids; as
salmon grow, certain fatty acid concentrations in their muscle tissue increase.
This relationship can be further visualized as follows.

```{r}
#| label: fatty_and_treatment_vis
#| fig-cap: Concentrations of selected fatty acids at the start and end of the trial, with an indication of whether animals received treatment.

p <- plotExpression(
    tse, rownames(tse), assay.type = "counts", x = "timepoint",
    colour_by = "study_group", scales = "free")
p
```

## Analyzing microbiota: Time and treatment effects

### Microbial composition

As a first step in analysing microbiota data, we summarize the microbial
composition with relative abundance barplot.

```{r}
#| label: abundance_plot
#| fig-cap: Relative abundance of core microbial genera across samples.

p <- plotAbundance(
    altExp(mae[[2]], "prev_genus"),
    assay.type = "relabundance",
    col.var = c("study_group", "timepoint"),
    facet.cols = TRUE, scales = "free_x"
    ) +
  guides(fill = guide_legend(title = "Genus"))
p
```

Salmon gut seems to be dominated by either genus _Mycoplasma_ or
_Photobacterium_.

### Association of alpha diversity with treatment and salmon age

Now, let’s proceed to calculate the Shannon alpha diversity index.

```{r}
#| label: alpha

# Calculate alpha diversity
mae[[2]] <- addAlpha(mae[[2]])
```

With the alpha diversity indices calculated and added to `colData`, we can now
assess whether time and treatment influence the microbial diversity in the
salmon gut flora.

```{r}
#| label: alpha2

# Get sample metadata
df <- colData(mae[[2]])
# Fit model to estimate influence of treatment and time to diversity
res <- lm(rank(shannon_diversity) ~ study_group + timepoint, data = df)
res <- summary(res)
res
```

Based on the results, we conclude that older salmon exhibit distinct microbial
diversity compared to younger ones. Additionally, there appears to be a
slight — though not statistically significant — effect of treatment on microbial
diversity.

```{r}
#| label: alpha3
#| fig-cap: Shannon diversity of microbial communities in salmon at the start and end of the trial, with information on treatment status.

p <- plotColData(
    mae[[2]], x = "timepoint", y = "shannon_diversity",
    colour_by = "study_group")
p
```

### Microbial dissimilarity among samples

Let's analyze whether we can find similar effect with beta diversity. Here we
perform Principal Coordinate Analysis (PCoA) with Bray-Curtis dissimilarity.

```{r}
#| label: pcoa_beta_diversity
#| fig-cap: PCoA (Bray-Curtis) of microbial data.

# Run PCoA
mae[[2]] <- runMDS(
    mae[[2]],
    FUN = getDissimilarity,
    method = "bray",
    assay.type = "relabundance"
)

# Display dissimilarity on a plot
p <- plotReducedDim(mae[[2]], "MDS", colour_by = "timepoint")
p
```

Distinct patterns in the PCoA plot show samples clustering by time points,
indicating that microbial profiles vary with salmon age. This reinforces the
results observed in alpha diversity, supporting an association between age and
shifts in microbial diversity.

## Multi-omics integration

Multi-omic factor analysis (MOFA) (see @Argelaguet2020) allows us to discover
latent factors that underlie the biological differences by taking in
consideration 2 or more omic assays. To cite the original authors, "MOFA can
be viewed as a statistically rigorous generalization of (sparse) principal
component analysis (PCA) to multi-omics data".

By applying MOFA analysis, our goal is to determine whether metagenomics and
fatty acids exhibit shared variability, ultimately assessing whether the
microbial community is associated with fatty acid composition.

```{r, results=FALSE}
#| label: fit_mofa
#| output: false

mae_temp <- mae
mae_temp[[2]] <- altExp(mae_temp[[2]], "prev_genus")

# Extract only transformed metagenomic assays for MOFA analysis
assays(mae_temp[[1]]) <- assays(mae_temp[[1]])[
    names(assays(mae_temp[[1]])) %in% c("standardize")
]
assays(mae_temp[[2]]) <- assays(mae_temp[[2]])[
    names(assays(mae_temp[[2]])) %in% c("counts")
]

# Transform MAE object to MOFA model
model <- create_mofa_from_MultiAssayExperiment(mae_temp)

# Set model's options
model_opts <- get_default_model_options(model)
model_opts$num_factors <- 5
model_opts$likelihoods[[2]] <- "poisson"
train_opts <- get_default_training_options(model)
train_opts$maxiter <- 20000

# Change convergence mode to slightly improve accuracy
train_opts$convergence_mode <- "slow"

# Prepare MOFA model
model <- prepare_mofa(
    object = model,
    model_options = model_opts,
    training_options = train_opts
)

# Train model
model <- run_mofa(model, use_basilisk = TRUE)
```

Next, we will plot the variances explained by each factor.

```{r}
#| label: var_factor1
#| fig-cap: Explained variance by the model for microbial and fatty acid data.

# Plot explained variances
p <- plot_variance_explained(model)
# Get explained variances from model as numeric values
df <- model@cache[[1]][[2]][[1]] |> stack()
df[["percentage"]] <- paste0(round(df[["value"]]), "%")
# Add them to plot
p <- p + geom_shadowtext(aes(label = df[["percentage"]]))
p
```

Factor 1 captures only variance within the metagenomics data, while over 2/3 of
variance captured by Factor 2 represents variance in fatty acids.
Factor 3 captures shared variability between the metagenomic data and
fatty acids, reflecting interconnected patterns between the two datasets.

Before exploring the shared variability, we first examine which metagenomic
variability is captured by Factor 1. We do not plot fatty acid weights because
the captured variability in factor 1 for fatty acids is 0% as seen in the
previous plot.

```{r}
#| label: plot_factor1
#| fig-cap: Features with the highest loadings for Factor 1.

p2 <- plot_top_weights(model, view = 2, factors = 1, nfeatures = 25) +
    labs(title = "Microbiota")

p2
```

From the plot above, we can see that the first factor captures mostly the
variability in _Mycoplasma_.

Let us then focus on loadings of factor 2.

```{r}
#| label: plot_factor2
#| fig-width: 10
#| fig-height: 6
#| fig-cap: Features with the highest loadings for Factor 2.

p1 <- plot_top_weights(model, view = 1, factors = 2, nfeatures = 25) +
    labs(title = "Fatty acids")
p2 <- plot_top_weights(model, view = 2, factors = 2, nfeatures = 25) +
    labs(title = "Microbiota")

p1 + p2
```

In the microbial data, particularly _Cetobacterium_, _Vibrio_, and _Aliivibrio_
show a negative association with Factor 2. Additionally, many fatty acids
display significant negative weights in this factor, though no single fatty
acid can be specifically tied to these taxa. This suggests that as the
abundances of these taxa is rise (or decrease), there is a corresponding
increase (or decrease) in overall fatty acid levels.

Next, we visualize Factor 3 that captured variance more evenly between microbes
and fatty acids.

```{r}
#| label: plot_factor3
#| fig-width: 10
#| fig-height: 6
#| fig-cap: Features with the highest loadings for Factors 3.

p1 <- plot_top_weights(model, view = 1, factors = 3, nfeatures = 25) +
    labs(title = "Fatty acids")
p2 <- plot_top_weights(model, view = 2, factors = 3, nfeatures = 25) +
    labs(title = "Microbiota")

p1 + p2
```

From the shared Factor 3, _Photobacterium_ emerge prominently. Similarly to
Factor 2, no single fatty acid can be directly associated with any of the
microbial species, including _Photobacterium_.

Worth noting is that, out of these 5 taxa, only _Mycoplasma_ does not appear to
share any variability with fatty acids as all its variability was captured by
the first factor which did not associate with fatty acids.

## Conclusions

The present case study has demonstrated how easy and fast it is to
download large dataset and transform the data into a `MultiAssayExperiment`,
which in turn gives the researchers access to an extensive plethora of
downstream tools, such `mia` and `MOFA2` that can be used to pre-process and
visualize the multi-omics data.

## Appendix

```{r}
#| label: create_summary_plot
#| fig-height: 6
#| fig-width: 12

features <- c(
    "Docosahexaenoic acid 22:6n-3 (DHA)", "Eicosapentaenoic acid 20:5n-3 (EPA)")
tse <- mae[[1]]
tse <- tse[features, ]
rownames(tse) <- c("DHA", "EPA")
p1 <- plotExpression(
    tse, rownames(tse), assay.type = "counts", x = "timepoint",
    colour_by = "study_group", scales = "free", ncol = 1) +
    labs(x = "Time point", y = "Concentration [mg/g]") +
    guides(colour = guide_legend(title = "Study group"))

p2 <- plotAbundance(
    altExp(mae[[2]], "prev_genus"),
    assay.type = "relabundance",
    col.var = c("study_group", "timepoint"),
    facet.cols = TRUE, scales = "free_x"
    ) +
    labs(y = "Relative abundance") +
  guides(fill = guide_legend(title = "Genus"))

p1 + p2  + plot_layout(widths = c(1, 2))
```

```{r session_info}
sessionInfo()
```
